Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-25 paper.tex
Fixed a few notes

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
\newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}

@@

\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}

@@

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and interactions
involved in architectures is a challenge as it often leads to complex
systems that are therefore prone to failure. To offer good user
experience, science gateways should remain as transparent as possible,
which requires robustness.
\note{Marc-Etienne}{I find that something is weak in this last
paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
Also, the text seem to timidly favor architectures with less components/interactions
without much explanation... Complex systems can be robust, but at greater costs...
Here is one:
"Architectures showing reduced system complexity will greatly facilitate the implementation
effort towards robust interactions which, in turn, have a positive impact on characteristics
such as gateway predictability, transparence, general reliability and results traceability and
reproducibility; characteristics that are also key to a good user experience."}

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.
\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
1- The interactions are padded in the middle of other stuff (Administrator, Data)
2- The Software component example says "Science gateway", should it not be empty?
3- The red arrow to explain the use of red; it looks like a very specific type
of interactions instead of the desired "any interactions or components can be red meaning...",
especially with 'a' and 'd' which the reader does not know the meaning yet.
Maybe a little red arrow and square with "Red used when specific to workflow execution"?
I don't have a great solution...
Red has special meaning, maybe remove it from the other pics (workflow eng, tasks)
as it creates lots of visual interference when looking for red?
4- Order in text jumps all over the fig notations. Explanations are given for, in order:
Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
}

@@

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc.

@@

\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/notations.pdf_tex}
\caption{Graphical notations}
\label{fig:notations}
\end{figure}

@@

\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \note{Marc-Etienne}{this -c2- can be also trigerred
    by the user}\note{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?} Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\note{sorina}{VIP and Moteur also share the WorkflowsDB}. 
In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment (DARE~\cite{maddineni2012distributed})
and in LONI Pipeline Environment~\cite{dinov2009efficient}.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
implemented in 3 different types of calls to the pool: workflow
submission (\texttt{1} \& \texttt{2}), workflow status retrieval
(\texttt{a} and \texttt{b})\note{sorina}{I suppose letters come here 
from the initial pool-model figure; it's a bit puzzling when mixed with 
letters from Fig 2}, and workflow retrieval (\texttt{13} \&
\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
through 2 types of calls: workflow instance retrieval (\texttt{4},
\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
and \texttt{12}). Workflow instance retrieval is used by the agents to
fetch work from the pool. Workflow instance update is used by the
agents to update workflow statuses. In the SHIWA pool, agents can wrap
different types of workflow engines to execute workflows expressed
with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
workflow engine plugins to declare their supported language and launch
engines, and by workflow engines to report their status to the
agent. These calls are specific to the SHIWA Pool implementation of
the \texttt{agent} component and therefore have no corresponding
representation in Figure~\subref*{archi:agent}.

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
\newcommand{\note}[2]{\textbf{\Large{\color{blue}\footnote{{\color{blue}\textbf{#1:} \textit{#2}\color{black}}}}}}
\newcommand{\closednote}[4]{}

@@

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}

@@

Architectures showing reduced system complexity will greatly
facilitate the implementation effort towards robust interactions
which, in turn, have a positive impact on characteristics such as
gateway predictability, transparence, general reliability and results
traceability and reproducibility; characteristics that are also key to
a good user experience. \closednote{Marc-Etienne}{I find that
  something is weak in this last paragraph. Robustness will help with
  transparency, sure, but it seems only a small facet.  Also, the text
  seem to timidly favor architectures with less
  components/interactions without much explanation... Complex systems
  can be robust, but at greater costs...  Here is one: "Architectures
  showing reduced system complexity will greatly facilitate the
  implementation effort towards robust interactions which, in turn,
  have a positive impact on characteristics such as gateway
  predictability, transparence, general reliability and results
  traceability and reproducibility; characteristics that are also key
  to a good user experience."}{Tristan}{Agreed.}

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.  \closednote{Marc-e}{A
  few things bother me a bit with the notation fig., making it harder
  to assimilate 1- The interactions are padded in the middle of other
  stuff (Administrator, Data) 2- The Software component example says
  "Science gateway", should it not be empty?  3- The red arrow to
  explain the use of red; it looks like a very specific type of
  interactions instead of the desired "any interactions or components
  can be red meaning...", especially with 'a' and 'd' which the reader
  does not know the meaning yet.  Maybe a little red arrow and square
  with "Red used when specific to workflow execution"?  I don't have a
  great solution...  Red has special meaning, maybe remove it from the
  other pics (workflow eng, tasks) as it creates lots of visual
  interference when looking for red?  4- Order in text jumps all over
  the fig notations. Explanations are given for, in order: Components,
  workflow engine, interactions, abstract interactions, red color,
  tasks, data (no Administrator?)...  Could re-ordering the fig fix
  this? A figure figure legend could go a long way too (are there some
  in this journal?)  }{Tristan}{Marc-E, is Figure 1 better now?}

@@

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc. Administrator refers to a science gateway user with
extended privileges to install applications.

@@

\begin{figure}
\centering
\def\svgwidth{0.7\columnwidth}
\input{figures/notations.pdf_tex}
\caption{Graphical notations}
\label{fig:notations}
\end{figure}

@@

\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?} Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment (DARE~\cite{maddineni2012distributed})
and in LONI Pipeline Environment~\cite{dinov2009efficient}.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In this
Figure, interaction \texttt{d} of Figure~\subref*{archi:agent} is
implemented in 3 different types of calls to the pool: workflow
submission (\texttt{1} \& \texttt{2}), workflow status retrieval
(\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come here 
from the initial pool-model figure; it's a bit puzzling when mixed with 
letters from Fig 2}{Tristan}{Fixed.}, and workflow retrieval (\texttt{13} \&
\texttt{14}). In Figure~\ref{fig:shiwa-pool-architecture},
interaction \texttt{f} of Figure~\subref*{archi:agent} is implemented
through 2 types of calls: workflow instance retrieval (\texttt{4},
\texttt{5} and \texttt{6}), and workflow instance update (\texttt{11}
and \texttt{12}). Workflow instance retrieval is used by the agents to
fetch work from the pool. Workflow instance update is used by the
agents to update workflow statuses. In the SHIWA pool, agents can wrap
different types of workflow engines to execute workflows expressed
with different languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and
\texttt{9} on Figure~\ref{fig:shiwa-pool-architecture} are used by
workflow engine plugins to declare their supported language and launch
engines, and by workflow engines to report their status to the
agent. These calls are specific to the SHIWA Pool implementation of
the \texttt{agent} component and therefore have no corresponding
representation in Figure~\subref*{archi:agent}.
